Large squeezing in Bose-Einstein condensates with large numbers of atoms

### 2013-07-16 large_number_squeezing_with_atoms.tex
First set of proofreading edits

Quantum degenerate ultracold atomic sources have been used to study fundamental processes such as quantum phase transitions \cite{Leggett2001,Buluta2009}, quantum non-equilibrium thermodynamics \cite{Kinoshita:2006,Barnett:2011} and entanglement of massive particles \cite{Kheruntsyan:2005}.  Their ability to be highly isolated from their environment and controlled precisely by optical, magnetic and rf fields makes them an ideal platform for engineering and manipulating non-trivial quantum states. In quantum optics, the most common way to produce entangled states has been through squeezing, and it has been recognised that the intrinsic nonlinearities of atoms naturally lead to quadrature squeezing \cite{Kheruntsyan:2005, johnssonET2007, haineET2009, doeringET2010}.  Number squeezing can be achieved with spin squeezing techniques \cite{kitagawaET1993, gross2010, sinatra2012}, molecular dissociation of Bose-Einstein Condensates (BECs) \cite{kheruntsyanET2002,davisET2008,savage2007,ogren2010}, four-wave mixing \cite{perrin2008, ogren2009, jascula2010}, double-well potentials \cite{esteve2008, maussang2010}, or mixing quadrature-squeezed states using Ramsey interferometry \cite{haineET2009}. The creation of these non-classical atomic states has relevance for fundamental tests of quantum mechanics such as the EPR paradox \cite{ogren2010, pu2000, sorensen2001, haine2005,zhao2007}, as well as atom interferometry \cite{doeringET2010, gross2010, sorensenET1999, kuzmichET2000, meyerET2001, liebfriedET2004, roosET2006}.
While these techniques have resulted in significant experimentally detectable number squeezing \cite{esteve2008,gross2010}, they have been limited to small numbers of atoms, typically a few hundred.

Ultracold atomic interferometers are also reaching state of the art sensitivities in precision measurements \cite{robins:2013}. While all current atom interferometry is limited by technical noise sources, it appears that BEC-based coherent sources offer the best chance of overcoming some of these issues, for the same reasons that optical lasers are typically preferred for optical interferometry \cite{szigeti:2012}. If these technical noise sources can be controlled, atom interferometers will be limited by atomic shot noise, which leads to a sensitivity that scales as the square root of the atomic flux.   From that point, sensitivity will only be improved by increasing the atomic flux, and/or using non-classical quantum states such as squeezed states to go below the shot noise limit.  Producing squeezed states for metrological purposes is therefore only relevant when it can be achieved in a context of large numbers of atoms.  This paper examines the generation of atomic squeezing using the inherent Kerr-like nonlinearities of ultracold atoms, and demonstrates a trap geometry in which metrologically relevant levels of squeezing are present in trap geometries that are compatible with large atomic number.

The creation, detection and application of squeezing requires precise mode-matching, and is therefore very sensitive to uncontrolled coupling between spatial modes of the atomic field.  Atomic squeezing experiments have typically attempted to operate in the single-mode regime, where the trap is tight enough such that all higher modes are frozen out \cite{LiET2009}.  However, tight traps limit the total number of atoms, as the 3-body recombination rate places an effective maximum atomic density in the gas, and well before that limit, the interatomic interactions break the single-mode operation.  Both of these effects limit the number of atoms that can be effectively squeezed, so we will examine a broader range of trap geometries, and calculate the dynamics in the presence of multiple spatial modes.
Many similar previous simulations have been performed in one or two spatial dimensions, by integrating out some of the remaining dimensions \cite{Kheruntsyan:2005, johnssonET2007, haineET2009, kheruntsyanET2002}.  For simulations of semiclassical fields this can be a reasonable approximation, but it can ignore spontaneous scattering processes in full quantum field theory calculations, so we take care to include full 3D simulations.

@@

\section{Model and two mode solutions}
\label{secTwoModeAnalytic}
Our goal is to describe the number squeezing possibilities of a BEC comprising two relevant internal states denoted by $|a\rangle$ and $|b\rangle$.  The Hamiltonian for a pair of coupled atomic internal modes is the spatial integral of the Hamiltonian density:
\begin{equation}
\hat{\mathcal{H}} = \sum_{j\in \{a,b\}} \hat{\psi}_j^{\dagger}\left(-\frac{\hbar^2}{2 m}\nabla^2+V_j\right)\hat{\psi}_j 
          + \sum_{i,j}\frac{U_{i j}}{2} \hat{\psi}_i^{\dagger} \hat{\psi}_j^{\dagger} \hat{\psi}_j \hat{\psi}_i
          + \kappa \hat{\psi}_a^{\dagger} \hat{\psi}_b + \kappa^* \hat{\psi}_b^{\dagger}  \hat{\psi}_a,
\label{eqFieldHamiltonian}
\end{equation}
where $\hat{\psi}_j(\mathbf{x})$ is the annihilation field operator for a particle at position $\mathbf{x}$ and in internal state $|j\rangle$, $m$ is the mass of the atoms, $V_j(\mathbf{x})$ is the trapping potential for atoms in internal state $|j\rangle$, $U_{ij}$ describe the various inter- and intra-state nonlinearities, and $\kappa$ describes the coupling between the two states.

@@

When considering squeezing, the Hamiltonian \ref{eqTwoModeHamiltonian} has typically been handled within the framework of spin squeezing, which has a long history \cite{kitagawaET1993, esteve2008,sorensenET1999, liebfriedET2004, wineland1994, steel+collett}. This approach involves exploiting the equivalence between the algebra of two harmonic oscillators and that of angular momentum, and writing
\begin{eqnarray}
\hat{S}_+ &=& \hat{a}^{\dagger} \hat{b} \\
\hat{S}_- &=& \hat{a} \hat{b}^{\dagger} \\
\hat{S}_x &=& \frac{1}{2} \left( \hat{S}_+ + \hat{S}_-\right) \\
\hat{S}_y &=& \frac{1}{2i} \left( \hat{S}_+ - \hat{S}_-\right) \\
\hat{S}_z &=& \frac{1}{2} \left( \hat{a}^{\dagger} \hat{a} - \hat{b}^{\dagger} \hat{b} \right).
\end{eqnarray}
Using these variables Eq.~(\ref{eqTwoModeHamiltonian}) is transformed into the Josephson Hamiltonian \cite{steel+collett}
\begin{equation}
\Delta \omega \, \hat{S}_z + \chi \hat{J}_z^2 + \Omega \hat{J}_x
\label{eqJosephson}
\end{equation}
where the $\Delta \omega$ is a function of the various nonlinearities and $\chi=\chi_{aa}+\chi_{bb}-2\chi{ab}$ are functions of the various nonlinearities, which is then solved using angular momentum algebra.

We take a different approach and solve Eq.~(\ref{eqTwoModeHamiltonian}) directly for two reasons. First, we want access to the number variances of each state, not just the variance of the number difference between the states. Second, Eq.~(\ref{eqJosephson}) has led to the common misconception that no squeezing can be generated in this system if $\chi_{aa}+\chi_{bb}-2\chi_{ab}=0$, as the effects of the nonlinearities on $\Delta \omega$ are dismissed as an irrelevent detuning. This is only the case if a spin-echo pulse is used half way through the application of (\ref{eqTwoModeHamiltonian}) (which is the protocol used in most spin squeezing experiments) or extra detunings are added to (\ref{eqJosephson}). Directly solving the Hamiltonian (\ref{eqTwoModeHamiltonian}) makes this clearer.

The system is prepared with all the population initially in $|a\rangle$, and vacuum in $|b\rangle$. We choose state $|a\rangle$ to initially be a coherent state with average number $\langle \hat{a}^{\dagger} \hat{a} \rangle = N$, where $N$ is the total particle number.  As the system is insensitive to the initial phase of the coherent state, this is equivalent to a mixture of coherent states of uncertain phase, or a mixture of Poissonian-distributed number states. Such a state is consistent with BEC coherence experiments \cite{Hadzibabic2004}.

At time $t=0$, the coupling $\Omega$ is applied until time $t=t_1$, resulting in a portion of the population being transferred into mode $|b\rangle$. We assume this transfer process is fast relative to the nonlinear energy time scale, ensuring that, as the initial states were coherent states, after the population transfer both $|a\rangle$ and $|b\rangle$ are still described by coherent states with mean number $\langle \hat{a}^{\dagger} \hat{a} \rangle = n_a$ and $\langle \hat{b}^{\dagger} \hat{b} \rangle = n_b$ respectively.  The coupling $\Omega$ is then switched off until time $t_2$, while the atoms interact solely through the nonlinear terms. After this hold time $\tau_{\mathrm{hold}} = t_2 - t_1 $, the coupling $\Omega$ is switched back on until time $t_3$, with a phase shift $\phi$ compared to its first application. During this last stage, the two modes exchange population and the quadrature variance is converted into number variance, in the same way that homodyne measurements are used in quantum optics to convert quadrature squeezing into number squeezing, which can be directly measured. In this interpretation, $\phi$ is the relative phase of the strong local oscillator, which allows specific phase angles of quadrature squeezing to be examined.  This experiment can also be interpreted as a Ramsey interferometer with a final beam splitter phase of $\phi$.  The entire sequence of pulses and the resulting populations are shown schematically in Figure~\ref{figPulseScheme}.

@@

We solve this system analytically, and derive expressions for absolute and relative number squeezing. We present the solutions here; the full derivation can be found in \ref{appendixTwoModeDerivation} as it is somewhat laborious. If we define
\begin{eqnarray}
s &=& \sin(\theta) \label{eqsDef} \\
c &=& \cos(\theta) \\
\theta &=& \Omega(t_3 - t_2) \\
\lambda_{ij} &=& \chi_{ij} \tau_{\mathrm{hold}} \\
A &=& \sqrt{n_a} \exp [n_a (e^{-i(\lambda_{aa}-\lambda_{ab})} - 1 )] \\
A_2 &=& n_a \exp [n_a (e^{-2i(\lambda_{aa}-\lambda_{ab})} - 1 )] \\
B &=& -i e^{i \phi}\sqrt{n_b} \exp [n_b (e^{-i(\lambda_{bb}-\lambda_{ab})} - 1 )] \\
B_2 &=& -n_b e^{2i\phi} \exp [n_b (e^{-2i(\lambda_{bb}-\lambda_{ab})} - 1 )] \\
D &=& AB^* - A^*B \label{eqDDef}
\end{eqnarray}
then the expectation values for number, number variance, and number difference variance after applying the final coupling field are given by
\begin{eqnarray}
N_a(t_3) &=& n_a c^2 + n_b s^2 + i c s D \\
%
N_b(t_3) &=& n_a s^2 + n_b c^2 - i c s D \\
%
{\mathrm{Var}} [ N_a](t_3) &=& n_a c^2 + n_b s^2 + 2 n_a n_b c^2 s^2 \nonumber\\
       && + c^2 s^2 (D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} B_2 A_2^* - e^{-i(\lambda_{aa} - \lambda_{bb})} B_2^* A_2) \nonumber \\
       && + i c s D (1-2 n_a c^2 -2 n_b s^2)  \nonumber\\
       && + 2 i c^3 s n_a (e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* - e^{i(\lambda_{aa} - \lambda_{ab})} A^* B ) \nonumber \\
       && + 2 i c s^3 n_b (e^{i(\lambda_{bb} - \lambda_{ab})} A B^* - e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B ) \\
%
{\mathrm{Var}} [ N_b](t_3) &=&  n_a s^2 + n_b c^2 + 2 n_a n_b c^2 s^2 \nonumber \\
       && + c^2 s^2 (D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} B_2 A_2^* - e^{-i(\lambda_{aa} - \lambda_{bb})} B_2^* A_2 ) \nonumber \\
       && + i c s D (-1+2 n_a s^2 + 2 n_b c^2)  \nonumber \\
       && + 2 i c^3 s n_b (e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B - e^{i(\lambda_{bb} - \lambda_{ab})} A B^* ) \nonumber \\
       && + 2 i c s^3 n_a (e^{i(\lambda_{aa} - \lambda_{ab})} A^* B - e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* ) \\
%
{\mathrm{Var}} [ N_a - N_b](t_3) &=& n_a + n_b + 4 i c s D (n_a - n_b)(s^2 - c^2) \nonumber \\
       && + 4 c^2 s^2 (2 n_a n_b +D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} A_2^* B_2 - e^{-i(\lambda_{aa} - \lambda_{bb})} A_2 B_2^* ) \nonumber \\
       && + 4 i c s (c^2 - s^2) \left( n_a e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* - n_a e^{i(\lambda_{aa} - \lambda_{ab})} A^* B \right. \nonumber \\
       && \,\,\,\,\, \left. + n_b e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B - n_b e^{i(\lambda_{bb} - \lambda_{ab})} A B^* \right) \label{eqNumDiffVariance}
\end{eqnarray}
In the case where all the nonlinearities are equal, such that $\chi_{aa} = \chi_{ab} = \chi_{bb}$, the number variances all reduce to those of a coherent state, e.g. ${\mathrm{Var}}[N_a]=N_a$, and no number squeezing is possible. It should also be noted that, although there are three independent nonlinearities $\chi_{ij}$ in the Hamiltonian (\ref{eqTwoModeHamiltonian}), the expressions for the number variances only depend on differences. This means that number squeezing is parameterised by only two independent quantities, for example $\chi_{aa}-\chi_{ab}$ and $\chi_{aa}-\chi_{bb}$.

@@

This two-mode analytic model allows arbitrarily good number squeezing, provided there are no limits to the number of particles in the system or to the hold time. The two-mode approximation can be valid in tight traps with very low particle number, and this result correctly predicts the squeezing demonstrated by Gross \textit{et al.} \cite{gross2010}. Unfortunately, for larger numbers of atoms in realistic traps there are more than two modes present, and multimode effects limit the level of squeezing that can be achieved.

@@

An analytic solution to the multimode, higher-dimensional problem is intractable, so we must use a numerical solution.  We use stochastic methods based on phase space representations to simulate the dynamics of the multimode quantum fields.  These methods alleviate the problem of directly simulating elements of the Hilbert space, the size of which increases exponentially with the number of spatial modes \cite{gardiner1991, steelET1998}.  Stochastic methods achieve this by finding a sufficiently well-behaved quasi-probability representation for the density matrix, which can then be simulated as the average behaviour of a number of low-dimensional samples.  These sample objects are the same size as the equivalent classical field.  In our case, the only tractable phase space representation is the functional Wigner representation \cite{steelET1998}, which is well behaved when the quantum field is approximately Gaussian, as it is for coherent and squeezed states.  Stochastic methods based on the functional Wigner representation have been used to model the behaviour of BECs and atom lasers \cite{johnssonET2007,dallET2009,dennisET2010}, as the Wigner representation allows the mapping of the system to a Fokker-Planck equation with positive semi-definite diffusion matrix (unlike the $P$-representation \cite{gardiner1991}), and does not suffer exponential path-weighting problems that would constrain it to very short simulation intervals (as would the positive-$P$ representation \cite{gardiner1991}).

The equation of motion for the density matrix is defined by the Hamiltonian in Eq.~(\ref{eqFieldHamiltonian}).  This defines the evolution of the functional Wigner distribution of the system, which has a one-to-one correspondence with the density matrix.  The equation of motion for the functional Wigner distribution contains derivatives of third order which we assume to be negligible.  This uncontrolled approximation is called the Truncated Wigner Approximation (TWA), and while it has been used widely on ultracold gases \cite{johnssonET2007,haineET2009,steelET1998,dennisET2010,johnssonHope:2007}, care must be taken that the simulation remains valid. For our simulations we checked the results of our simulations against both the analytic two-mode solution, as well as checking for typical indications of TWA breakdown such as the appearance of negative densities in lightly populated modes.  We could also use the Bogoliubov analysis described in section \ref{sec:2D3Dsqueezing} as an independent estimate of the validity of the TWA.  We found that the TWA remained valid except in some 3D simulations with high nonlinearities, where there were a large number of modes, and the number of particles per mode could drop to ten or less in the densest regions where the squeezing was generated. Over very long time scales, this low mode occupation began to result in TWA breakdown.  Fortunately, the best squeezing was typically found in regions where the TWA was valid.

@@

These stochastic equations take the form
\begin{eqnarray}
\frac{d \phi_a}{dt} = -\frac{i}{\hbar}\left(-\frac{\hbar^2}{2 m}\nabla^2+V_a(\mathbf{x}) + U_{a a} n_a + U_{a b} \tilde{n}_b  \right)\phi_a   + \kappa \phi_b  \nonumber \\
\frac{d \phi_b}{dt} = -\frac{i}{\hbar}\left(-\frac{\hbar^2}{2 m}\nabla^2+V_b(\mathbf{x}) + U_{b b} n_b + U_{a b} \tilde{n}_a   \right)\phi_b   + \kappa^* \phi_a
  \label{eqStochasticEquations}
\end{eqnarray}
where $n_j=\left|\phi_{j}\right|^2 -\frac{1}{dV}$, $\tilde{n}_j=\left|\phi_{j}\right|^2 -\frac{1}{2 dV}$, $dV$ is the volume element of the grid on which the simulation is carried out, and the terms proportional to $1/dV$ correspond to vacuum corrections. It is a peculiarity of the Wigner representation that, up to the vacuum correction terms, Eqs.~(\ref{eqStochasticEquations}) look identical to those of the coupled semi-classical nonlinear Schr{\"{o}}dinger equation. The quantum statistical nature of the Wigner equations enters due to the fact that Eqs.~(\ref{eqStochasticEquations}) are run many times and stochastically averaged, with each run using different set of random initial conditions describing (in this case) the noise on a coherent state.

>>>>

Quantum degenerate ultracold atomic sources have been used to study fundamental processes such as quantum phase transitions \cite{Leggett2001,Buluta2009}, quantum non-equilibrium thermodynamics \cite{Kinoshita:2006,Barnett:2011} and entanglement of massive particles \cite{Kheruntsyan:2005}.  Their ability to be highly isolated from their environment and controlled precisely by optical, magnetic and rf fields makes them an ideal platform for engineering and manipulating non-trivial quantum states. In quantum optics, the most common way to produce entangled states has been through squeezing, and it has been recognised that the intrinsic nonlinearities of atoms naturally lead to quadrature squeezing \cite{Kheruntsyan:2005, johnssonET2007, haineET2009, doeringET2010}.  Number squeezing can be achieved with spin squeezing techniques \cite{kitagawaET1993, gross2010, sinatra2012}, molecular dissociation of Bose-Einstein Condensates (BECs) \cite{kheruntsyanET2002,davisET2008,savage2007,ogren2010}, four-wave mixing \cite{perrin2008, ogren2009, jascula2010}, double-well potentials \cite{esteve2008, maussang2010}, or mixing quadrature-squeezed states using Ramsey interferometry \cite{haineET2009}. The creation of these non-classical atomic states has relevance for fundamental tests of quantum mechanics such as the EPR paradox \cite{ogren2010, pu2000, sorensen2001, haine2005,zhao2007}, as well as atom interferometry \cite{doeringET2010, gross2010, sorensenET1999, kuzmichET2000, meyerET2001, liebfriedET2004, roosET2006}.
While these techniques have resulted in significant experimentally detectable number squeezing, they have been limited to small numbers of atoms, typically a few hundred \cite{gross2010, esteve2008}.

Ultracold atomic interferometers are also reaching state of the art sensitivities in precision measurements \cite{robins:2013}. While all current atom interferometry is limited by technical noise sources, it appears that BEC-based coherent sources offer the best chance of overcoming some of these issues, for the same reasons that optical lasers are typically preferred for optical interferometry \cite{szigeti:2012}. If these technical noise sources can be controlled, atom interferometers will be limited by atomic shot noise, which leads to a sensitivity that scales as the square root of the atomic flux.   From that point, sensitivity will only be improved by increasing the atomic flux, and/or using non-classical quantum states such as squeezed states to go below the shot noise limit \cite{robins:2013}.  Producing squeezed states for metrological purposes is therefore only relevant when it can be achieved in a context of large numbers of atoms.  This paper examines the generation of atomic squeezing using the inherent Kerr-like nonlinearities of ultracold atoms, and demonstrates a trap geometry in which metrologically relevant levels of squeezing are present in trap geometries that are compatible with large atomic number.

The creation, detection and application of squeezing requires precise mode-matching, and is therefore very sensitive to uncontrolled coupling between spatial modes of the atomic field.  Atomic squeezing experiments have typically attempted to operate in the single-mode regime, where the trap is tight enough such that all higher modes are frozen out \cite{LiET2009}.  However, tight traps limit the total number of atoms, as the 3-body recombination rate places an effective maximum atomic density in the gas, and well before that limit the interatomic interactions break the single-mode operation.  Both of these effects limit the number of atoms that can be effectively squeezed, so we will examine a broader range of trap geometries, and calculate the dynamics in the presence of multiple spatial modes.
Many similar previous simulations have been performed in one or two spatial dimensions, by integrating out some of the remaining dimensions \cite{Kheruntsyan:2005, johnssonET2007, haineET2009, kheruntsyanET2002}.  For simulations of semiclassical fields this can be a reasonable approximation, but it can ignore spontaneous scattering processes in full quantum field theory calculations, so we take care to include full 3D simulations.

@@

\section{Model and two mode solutions}
\label{secTwoModeAnalytic}
Our goal is to describe the number squeezing possibilities of a BEC comprising two relevant internal states denoted by $|a\rangle$ and $|b\rangle$.  The Hamiltonian for a pair of coupled atomic internal modes is the spatial integral of the Hamiltonian density:
\begin{equation}
\hat{\mathcal{H}} = \sum_{j\in \{a,b\}} \hat{\psi}_j^{\dagger}\left(-\frac{\hbar^2}{2 m}\nabla^2+V_j\right)\hat{\psi}_j 
          + \sum_{i,j}\frac{U_{i j}}{2} \hat{\psi}_i^{\dagger} \hat{\psi}_j^{\dagger} \hat{\psi}_j \hat{\psi}_i
          + \kappa \hat{\psi}_a^{\dagger} \hat{\psi}_b + \kappa^* \hat{\psi}_b^{\dagger}  \hat{\psi}_a,
\label{eqFieldHamiltonian}
\end{equation}
where $\hat{\psi}_j(\mathbf{r})$ is the annihilation field operator for a particle at position $\mathbf{r}$ and in internal state $|j\rangle$, $m$ is the mass of the atoms, $V_j(\mathbf{r})$ is the trapping potential for atoms in internal state $|j\rangle$, $U_{ij}$ describe the various inter- and intra-state nonlinearities, and $\kappa$ describes the coupling between the two states.

@@

When considering squeezing, the Hamiltonian (\ref{eqTwoModeHamiltonian}) has typically been handled within the framework of spin squeezing, which has a long history \cite{kitagawaET1993, esteve2008,sorensenET1999, liebfriedET2004, wineland1994, steel+collett}. This approach involves exploiting the equivalence between the algebra of two harmonic oscillators and that of angular momentum, and writing
\begin{eqnarray}
\hat{S}_+ &=& \hat{a}^{\dagger} \hat{b} \\
\hat{S}_- &=& \hat{a} \hat{b}^{\dagger} \\
\hat{S}_x &=& \frac{1}{2} \left( \hat{S}_+ + \hat{S}_-\right) \\
\hat{S}_y &=& \frac{1}{2i} \left( \hat{S}_+ - \hat{S}_-\right) \\
\hat{S}_z &=& \frac{1}{2} \left( \hat{a}^{\dagger} \hat{a} - \hat{b}^{\dagger} \hat{b} \right).
\end{eqnarray}
Using these variables Eq.~(\ref{eqTwoModeHamiltonian}) is transformed into the Josephson Hamiltonian \cite{steel+collett}
\begin{equation}
\hat{H}_J = \Delta \omega \, \hat{S}_z + \chi \hat{J}_z^2 + \Omega \hat{J}_x
\label{eqJosephson}
\end{equation}
where $\Delta \omega$ is a function of the various nonlinearities and $\chi=\chi_{aa}+\chi_{bb}-2\chi_{ab}$. The system described by $\hat{H}_J$ is then solved by using angular momentum algebra.

We take a different approach and solve Eq.~(\ref{eqTwoModeHamiltonian}) directly for two reasons. First, we want access to the number variances of each state, not just the variance of the number difference between the states. Second, Eq.~(\ref{eqJosephson}) has led to the common misconception that no squeezing can be generated in this system if $\chi_{aa}+\chi_{bb}-2\chi_{ab}=0$, as the effects of the nonlinearities on $\Delta \omega$ are dismissed as an irrelevent detuning. This is only the case if a spin-echo pulse is used half way through the application of (\ref{eqTwoModeHamiltonian}) (which is the protocol used in most spin squeezing experiments) or extra detunings are added to (\ref{eqJosephson}). Directly solving Eq.~(\ref{eqTwoModeHamiltonian}) makes this clearer.

The system is prepared with all the population initially in $|a\rangle$, and vacuum in $|b\rangle$. We choose the initial state $|a\rangle$ to be a coherent state with average number $\langle \hat{a}^{\dagger} \hat{a} \rangle = N$, where $N$ is the total particle number.  As the system is insensitive to the initial phase of the coherent state, this is equivalent to a mixture of coherent states of uncertain phase, or a mixture of Poissonian-distributed number states. Such a state is consistent with BEC coherence experiments \cite{Hadzibabic2004}.

At time $t=0$, the coupling $\Omega$ is applied until time $t=t_1$, resulting in a portion of the population being transferred into mode $|b\rangle$. We assume this transfer process is fast relative to the nonlinear energy time scale, ensuring that, as the initial states were coherent states, after the population transfer both $|a\rangle$ and $|b\rangle$ are still described by coherent states with mean number $\langle \hat{a}^{\dagger} \hat{a} \rangle = n_a$ and $\langle \hat{b}^{\dagger} \hat{b} \rangle = n_b$ respectively.  The coupling $\Omega$ is then switched off until time $t_2$, while the atoms interact solely through the nonlinear terms. After this hold time $\tau_{\mathrm{hold}} = t_2 - t_1 $, the coupling $\Omega$ is switched back on until time $t_3$, with a phase shift $\phi$ compared to its first application. During this last stage, the two modes exchange population and the quadrature variances are converted into number variance, in the same way that homodyne measurements are used in quantum optics to convert quadrature squeezing into number squeezing, which can be directly measured. In this interpretation, $\phi$ is the relative phase of the strong local oscillator, which allows specific phase angles of the quadrature squeezing to be examined.  This experiment can also be interpreted as a Ramsey interferometer with a final beam splitter phase of $\phi$.  The entire sequence of pulses and the resulting populations are shown schematically in Figure~\ref{figPulseScheme}.

@@

We solve this system analytically, and derive expressions for absolute and relative number squeezing. We present the solutions here; the full derivation can be found in \ref{appendixTwoModeDerivation} as it is somewhat laborious. If we define
\begin{eqnarray}
s &=& \sin(\theta) \label{eqsDef} \\
c &=& \cos(\theta) \\
\theta &=& \Omega(t_3 - t_2) \\
\lambda_{ij} &=& \chi_{ij} \tau_{\mathrm{hold}} \\
A &=& \sqrt{n_a} \exp [n_a (e^{-i(\lambda_{aa}-\lambda_{ab})} - 1 )] \\
A_2 &=& n_a \exp [n_a (e^{-2i(\lambda_{aa}-\lambda_{ab})} - 1 )] \\
B &=& -i e^{i \phi}\sqrt{n_b} \exp [n_b (e^{-i(\lambda_{bb}-\lambda_{ab})} - 1 )] \\
B_2 &=& -n_b e^{2i\phi} \exp [n_b (e^{-2i(\lambda_{bb}-\lambda_{ab})} - 1 )] \\
D &=& AB^* - A^*B \label{eqDDef}
\end{eqnarray}
then the expectation values for number, number variance, and number difference variance after applying the final coupling field are given by
\begin{eqnarray}
N_a(t_3) &=& n_a c^2 + n_b s^2 + i c s D \\
%
N_b(t_3) &=& n_a s^2 + n_b c^2 - i c s D \\
%
{\mathrm{Var}} [ N_a](t_3) &=& n_a c^2 + n_b s^2 + 2 n_a n_b c^2 s^2 \nonumber\\
       && + c^2 s^2 (D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} B_2 A_2^* - e^{-i(\lambda_{aa} - \lambda_{bb})} B_2^* A_2) \nonumber \\
       && + i c s D (1-2 n_a c^2 -2 n_b s^2)  \nonumber\\
       && + 2 i c^3 s n_a (e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* - e^{i(\lambda_{aa} - \lambda_{ab})} A^* B ) \nonumber \\
       && + 2 i c s^3 n_b (e^{i(\lambda_{bb} - \lambda_{ab})} A B^* - e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B ) \\
%
{\mathrm{Var}} [ N_b](t_3) &=&  n_a s^2 + n_b c^2 + 2 n_a n_b c^2 s^2 \nonumber \\
       && + c^2 s^2 (D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} B_2 A_2^* - e^{-i(\lambda_{aa} - \lambda_{bb})} B_2^* A_2 ) \nonumber \\
       && + i c s D (-1+2 n_a s^2 + 2 n_b c^2)  \nonumber \\
       && + 2 i c^3 s n_b (e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B - e^{i(\lambda_{bb} - \lambda_{ab})} A B^* ) \nonumber \\
       && + 2 i c s^3 n_a (e^{i(\lambda_{aa} - \lambda_{ab})} A^* B - e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* ) \\
%
{\mathrm{Var}} [ N_a - N_b](t_3) &=& n_a + n_b + 4 i c s D (n_a - n_b)(s^2 - c^2) \nonumber \\
       && + 4 c^2 s^2 (2 n_a n_b +D^2 - e^{i(\lambda_{aa} - \lambda_{bb})} A_2^* B_2 - e^{-i(\lambda_{aa} - \lambda_{bb})} A_2 B_2^* ) \nonumber \\
       && + 4 i c s (c^2 - s^2) \left( n_a e^{-i(\lambda_{aa} - \lambda_{ab})} A B^* - n_a e^{i(\lambda_{aa} - \lambda_{ab})} A^* B \right. \nonumber \\
       && \,\,\,\,\, \left. + n_b e^{-i(\lambda_{bb} - \lambda_{ab})} A^* B - n_b e^{i(\lambda_{bb} - \lambda_{ab})} A B^* \right) \label{eqNumDiffVariance}
\end{eqnarray}
In the case where all the nonlinearities are equal, such that $\chi_{aa} = \chi_{ab} = \chi_{bb}$, the number variances all reduce to those of a coherent state, i.e. ${\mathrm{Var}}[N_a]=N_a$, and no number squeezing is possible. It should also be noted that, although there are three independent nonlinearities $\chi_{ij}$ in the Hamiltonian (\ref{eqTwoModeHamiltonian}), the expressions for the number variances only depend on differences. This means that number squeezing is parameterised by only two independent quantities, for example $\chi_{aa}-\chi_{ab}$ and $\chi_{aa}-\chi_{bb}$.

@@

This two-mode analytic model allows arbitrarily good number squeezing, provided there are no limits to the number of particles in the system or to the hold time. The two-mode approximation can be valid in tight traps with very low particle number, and this result correctly predicts the squeezing demonstrated by Gross \textit{et al.} \cite{gross2010}. Unfortunately, for larger numbers of atoms in realistic traps there are more than two modes present, and these multimode effects limit the level of squeezing that can be achieved.

@@

An analytic solution to the multimode, higher-dimensional problem is intractable, so we must turn to numerical solutions.  We use stochastic methods based on phase space representations to simulate the dynamics of the multimode quantum fields.  These methods alleviate the problem of directly simulating elements of the Hilbert space, the size of which increases exponentially with the number of spatial modes \cite{gardiner1991, steelET1998}.  Stochastic methods achieve this by finding a sufficiently well-behaved quasi-probability representation for the density matrix, which can then be simulated as the average behaviour of a number of low-dimensional samples.  These sample objects are the same size as the equivalent classical field.  In our case, the only tractable phase space representation is the functional Wigner representation \cite{steelET1998}, which is well behaved when the quantum field is approximately Gaussian, as it is for coherent and squeezed states.  Stochastic methods based on the functional Wigner representation have been used to model the behaviour of BECs and atom lasers \cite{johnssonET2007,dallET2009,dennisET2010}, as the Wigner representation allows the mapping of the system to a Fokker-Planck equation with positive semi-definite diffusion matrix (unlike the $P$-representation \cite{gardiner1991}), and does not suffer exponential path-weighting problems that would constrain it to very short simulation intervals (as would the positive-$P$ representation \cite{gardiner1991}).

The equation of motion for the density matrix is defined by the Hamiltonian in Eq.~(\ref{eqFieldHamiltonian}).  This defines the evolution of the functional Wigner distribution of the system, which has a one-to-one correspondence with the density matrix.  The equation of motion for the functional Wigner distribution contains derivatives of third order which we assume to be negligible.  This uncontrolled approximation is called the Truncated Wigner Approximation (TWA), and while it has been used widely on ultracold gases \cite{johnssonET2007,haineET2009,steelET1998,dennisET2010,johnssonHope:2007}, care must be taken that the simulation remains valid. For our simulations we checked the results of our simulations against the analytic two-mode solution, as well as checking for typical indications of TWA breakdown such as the appearance of negative densities in lightly populated modes.  We were also able to use the Bogoliubov analysis described in Section \ref{sec:2D3Dsqueezing} as an independent estimate of the validity of the TWA.  We found that the TWA remained valid except in some 3D simulations with high nonlinearities, where there were a large number of modes, and the number of particles per mode could drop to ten or less in the densest regions where the squeezing was generated. Over very long time scales, this low mode occupation began to result in TWA breakdown.  Fortunately, the best squeezing was typically found in regions where the TWA was valid.

@@

These stochastic equations take the form
\begin{eqnarray}
\frac{d \phi_a}{dt} = -\frac{i}{\hbar}\left(-\frac{\hbar^2}{2 m}\nabla^2+V_a(\mathbf{r}) + (U_{a a} n_a + U_{a b} \tilde{n}_b  \right)\phi_a   + \kappa \phi_b  \nonumber \\
\frac{d \phi_b}{dt} = -\frac{i}{\hbar}\left(-\frac{\hbar^2}{2 m}\nabla^2+V_b(\mathbf{r}) + U_{b b} n_b + U_{a b} \tilde{n}_a   \right)\phi_b   + \kappa^* \phi_a
  \label{eqStochasticEquations}
\end{eqnarray}
where $n_j=\left|\phi_{j}\right|^2 -\frac{1}{dV}$, $\tilde{n}_j=\left|\phi_{j}\right|^2 -\frac{1}{2 dV}$, $dV$ is the volume element of the grid on which the simulation is carried out, and the terms proportional to $1/dV$ correspond to vacuum corrections. It is a peculiarity of the Wigner representation that, up to the vacuum correction terms, Eqs.~(\ref{eqStochasticEquations}) look identical to those of the coupled semi-classical nonlinear Schr{\"{o}}dinger equation. The quantum statistical nature of the Wigner equations enters due to the fact that Eqs.~(\ref{eqStochasticEquations}) are run many times and stochastically averaged, with each run using different set of random initial conditions describing (in this case) the noise on a coherent state.

We integrate Eqs.~(\ref{eqStochasticEquations}) numerically in one, two and three spatial dimensions using the numerical package XMDS2 \cite{dennis2013}.  For the one- and two-dimensional simulations, the dimensional reduction is achieved by estimating the mode shape in three dimensions, and using dimensionally reduced values for $U_{ij}$ that match the zero-dimensional reduction given by Eq.~(\ref{eqChiUequivalence}).  To make relevant comparisons between equivalent situations with different numbers of spatial dimensions, this mode shape is typically chosen to match the chemical potentials of the initial states.
